---
title: "Heatmaps of gene exchange experiment"
date: '2022-11-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache.extra = 1,echo = F) 
library(dplyr)
library(ggplot2)
library(plotly)
#library(d3heatmap)
library(viridis)
library(hrbrthemes)
library(stringr)
library(heatmaply)
source('./R/plot.func.addition.R')

```

## Credits

-   Analysis by Chrispin Chaguza
-   Experiments by Anna York
-   Sequencing by Jason Rosch and colleagues (St Jude's)
-   Interactive plots by Dan Weinberger
-   Funding: Merck Investigator Studies Program

## Summary

These plots represent pangenomes created based on the sequenced strains. The heatmaps indicate whether a particular strain has **gained a gene (yellow)** or **lost a gene(dark blue)** compared to the reference. The plots at the bottom show the genomic location of genes gained and lost, mapped to a TIGR4 reference

```{r}
# key0 <-  read.table(file = './Data/Strain.clone.info.tsv', sep = '\t', header = TRUE) %>% 
#   dplyr::rename(parent_ST=ID, new_ST=Strain , strain=Hartwell)

key1 <-  readxl::read_excel('./Data/PathogenWatch/Pathogen Watch Output (Capsule Switch Mutants).xlsx') %>% 
  dplyr::rename(straininfo=`Capsule Switch Isolate`,strain=`Genome Name`) %>%
  mutate(parent_ST= sub("::.*", "", straininfo),
         parent_ST=  gsub("∆cps", "", parent_ST),
         new_ST= sub(".*::(\\S+) .*", "\\1", straininfo),
         Clone=extracted_text <- sub("^[^ ]+ (.*)", "\\1", straininfo,
)
)%>%
  dplyr::select(strain,parent_ST,new_ST,Clone)

g0 <- read.table(file = './Data/Pangenome.gene.presence.absence.tsv', sep = '\t', header = TRUE) %>%
  mutate(Cps.gene.name = if_else(is.na(Cps.gene.name), Gene.cluster, Cps.gene.name), 
         gene_name_cluster= paste(Cps.gene.name,Gene.cluster, sep='_'),
    gene_cluster_no_numb = ifelse(grepl('_', Gene.cluster), str_extract(Gene.cluster,  "^(.*?)(?=_)"), Gene.cluster
    )
    )%>%
  group_by(gene_cluster_no_numb) %>%
  arrange(Gene.cluster) %>%
  mutate(index=row_number(), 
         Gene.cluster2=paste(gene_cluster_no_numb, index, sep='_')) %>%
  ungroup()   
 

g2.labs <- key1$strain[key1$parent_ST=='15B/C']

g3.labs <- key1$strain[key1$parent_ST=='22F' & key1$strain != 'AYCs074'] #excludes 11A clone 1--this is an extreme outlier in terms of what is added in


# g1 <- g0 %>%
#   dplyr::select( gene_name_cluster, all_of(intersect(g1.labs, names(g0)))) %>%
#   reshape2::melt(., id.vars=c('gene_name_cluster')) %>%
#   reshape2::dcast(variable ~ gene_name_cluster) 
# 
# row.names(g1) <- g1$variable
# g1 <- g1 %>%   select(-variable)
# 
# prev_acq_g1 <- apply(g1, 2,mean)
# g1 <- g1[,prev_acq_g1>0 & prev_acq_g1<1]

##same for g2

g2 <- g0 %>%
  dplyr::select( Gene.cluster2, all_of(intersect(g2.labs, names(g0)))) %>%
  reshape2::melt(., id.vars=c('Gene.cluster2')) %>%
  reshape2::dcast(variable ~ Gene.cluster2) 

row.names(g2) <- g2$variable
g2 <- g2 %>%   select(-variable)

prev_acq_g2 <- apply(g2, 2,mean)
g2 <- g2[,prev_acq_g2>0.00 & prev_acq_g2<1.0]

##same for g3

g3 <- g0 %>%
  dplyr::select( Gene.cluster2, all_of(intersect(g3.labs, names(g0)))) %>%
  reshape2::melt(., id.vars=c('Gene.cluster2')) %>%
  reshape2::dcast(variable ~ Gene.cluster2) 

row.names(g3) <- g3$variable
g3 <- g3 %>%   select(-variable)

prev_acq_g3 <- apply(g3, 2,mean)
g3 <- g3[,prev_acq_g3>0.0 & prev_acq_g3<1.0]

a1 <- read.table(file = './Data/Genes.with.annotations.tsv', sep = '\t', header = TRUE) %>%
  dplyr::rename(Gene=Gene.cluster) %>%
  mutate(Gene= gsub('~~~~~~','......', fixed=T, Gene ),
         Non.unique.Gene.name = gsub(';','',Non.unique.Gene.name,fixed=T),
    Non.unique.Gene.name = if_else(Non.unique.Gene.name=='', Gene,Non.unique.Gene.name )) %>%
  #filter(Non.unique.Gene.name!='') %>%
  arrange(Non.unique.Gene.name) %>%
  group_by(Non.unique.Gene.name) %>%
  mutate(Gene.cluster2 = paste0(Non.unique.Gene.name,'_', row_number()),
        Cps.Gene = if_else(Cps.gene.name!='',1,0),
      gene_name_cluster= paste(Gene, Non.unique.Gene.name, sep='_')
) %>%
  ungroup() %>%
  dplyr::select(Gene.cluster2,Cps.Gene,gene_name_cluster, Gene,Non.unique.Gene.name,Cps.gene.name ,Sequence)


p2 <- plot.func.addition(g2)

p3 <- plot.func.addition(g3)


```

## Serotype 9V background

Parent strain is Serotype 9V ID ABC010001831, obtained from the CDC

```{r, fig.width=10, fig.height=5}
#p1$p
```

## Serotype 15B/C background

Parent strain is 15B/C, a carriage isolate from Hungary, ID Pool_hung_H10 (H124)

```{r, fig.width=10, fig.height=5}
p2$p
```

## Serotype 22F background

Parent strain is serotype 22F, ID: ABC010003966, obtained from the CDC (IPD isolate)

```{r, fig.width=10, fig.height=5}
p3$p
```

```{r}
#p1$gain_loss <- p1$gain_loss %>%
 # mutate(parentST='9V')

p2$gain_loss <- p2$gain_loss %>%
  mutate(parentST='15BC')

p3$gain_loss <- p3$gain_loss %>%
  mutate(parentST='22F')
  

#here we are looking for strains that have gain of one allele and loss of another..see 15B.C #3; gpsB_1
combined <- bind_rows( p2$gain_loss, p3$gain_loss) %>%
  filter(value %in% c(-1,1)) %>%
  arrange(Var1, Var2)
```

```{r, eval=F}
#prepare  for BLAST

library (devtools)
library (tidyverse)
#install.packages("BiocManager")
#BiocManager::install("Biostrings")
library(Biostrings)
source_url("https://raw.githubusercontent.com/lrjoshi/FastaTabular/master/fasta_and_tabular.R")

 a1 %>% select(allele, Sequence) %>%
  write.csv('./Data/genes.csv', row.names = F)
#TabularToFasta("./Data/genes.csv")
 
 source_url("https://raw.githubusercontent.com/lrjoshi/FastaTabular/master/fasta_and_tabular.R")


 prot1 <- a1 %>% select(allele, Sequence)

 prot2 <-  apply(prot1,1, function(x){
      as.character(translate(DNAString(x['Sequence'])))
      }, simplify=F) %>%
   unlist()

 
 prot3 <- cbind.data.frame('allele'=a1$allele, 'aa'=prot2 ) %>% 
   mutate(allele=paste0('gene_',allele)) %>%   
   write.csv('./Data/aa.csv', row.names = F)

TabularToFasta("./Data/aa.csv")  #note this generates a FASTA file on the home directory called dna_fasta.fasta

#IN tblastn, upload AA sequence, BLAST against TIGR4 specifically, then 'download all' as hit table csv. note use tblastn NOT blastn
```

## Genomic context of gain/loss based on mapping to TIGR4 genome

Set 1: Parent Strain is 9V

8, 7F, 7C, 6C, 22F, 19A
```{r}
loc1 <- read.csv('./Data/DBB51US6016-Alignment-HitTable.csv', header=F) %>%
  filter(V2=='CP089948.1') %>%
  dplyr::rename(allele1=V1, pct_ident=V3, start=V9, end=V10, positives_pct=V13, matchNaa=V4, query_length =V8) %>%
  mutate(pct_coverage = (matchNaa)/query_length*100
         ) %>%
  filter(pct_ident>80 & pct_coverage >80) %>%
  select(allele1, start, end,pct_ident,pct_coverage,query_length) %>%
  mutate(allele1 = gsub('gene_','',allele1),
           allele=str_extract(allele1, "^(.*?)(?=_)")) %>%
  group_by(allele) %>%
  mutate(allele_order=row_number()) %>%
  filter(allele_order==1) %>%
  ungroup()
# 
# p1a <- p1$gain_loss %>%
#   dplyr::rename(strain=Var1, allele=Var2) %>%
#   left_join(loc1, by='allele') %>%
#   filter(value %in% c(-1,1) & !is.na(start)) %>%
#   mutate(value=as.factor(value) ,strainN= as.numeric(as.factor(strain))+rnorm(n(), 0, 0.05) )
# 
# p2a <- ggplot(p1a, aes( y=strainN, x=start ,color=value,text=paste(strain,allele), alpha=0.5 )) +
#          geom_point(size=0.3) +
#        theme_classic() +
#     xlim(range(loc1$start) ) +
# 
#   geom_vline(xintercept=c(315597, 339358)) #roughly dexB and aliA
# 
# ggplotly(p2a)
```

Set 2: Parent strain is 15B/C, a carriage isolate from Hungary, ID Pool_hung_H10 (H124)


7F, 6C (2), 35B, 35F (2), 3, 22F, 16F, 15B/C, 11A have the cassette added around 1235484
```{r}

p1a <- p2$gain_loss %>%
  dplyr::rename(strain=Var1, allele1=Var2) %>%
  mutate(allele=str_extract(allele1, "^(.*?)(?=_)"))%>%
  left_join(loc1, by='allele') %>%
  filter(value %in% c(-1,1) & !is.na(start)) %>%
  mutate(value=as.factor(value) ,strainN= as.numeric(as.factor(strain))+rnorm(n(), 0, 0.05) )

p2a <- ggplot(p1a, aes( y=strainN, x=start ,color=value,text=paste(strain,allele), alpha=0.5 )) +
         geom_point(size=0.3) +
       theme_classic() +
    xlim(range(loc1$start) ) +
  geom_vline(xintercept=c(315597, 339358)) #roughly dexB and aliA

ggplotly(p2a)
```

Set 3: Parent strain is serotype 22F, ID: ABC010003966, obtained from the CDC (IPD isolate)

7F, 6C, 35B, 35F, 3, 22F,16F, 15B.C, 11A (this is the same set of serotypes seen for set 2...are these mislabeled cps genes? cps adjacent genes on these backgrounds? on set 1, some of the serotypes overlap, some do not.
```{r}

p1a <- p3$gain_loss %>%
  dplyr::rename(strain=Var1, allele1=Var2) %>%
    mutate(allele=str_extract(allele1, "^(.*?)(?=_)"))%>%
  left_join(loc1, by='allele') %>%
  filter(value %in% c(-1,1) & !is.na(start)) %>%
  mutate(value=as.factor(value) ,strainN= as.numeric(as.factor(strain))+rnorm(n(), 0, 0.05) )

p2a <- ggplot(p1a, aes( y=strainN, x=start ,color=value,text=paste(strain,allele), alpha=0.5 )) +
         geom_point(size=0.3) +
       theme_classic() +
    xlim(range(loc1$start) ) +
  geom_vline(xintercept=c(315597, 339358)) #roughly dexB and aliA

ggplotly(p2a)
```


## Same but align to Hungary-19A

```{r}
loc2 <- read.csv('./Data/hungary19A-tblastn-Alignment-HitTable.csv', header=F) %>%
  dplyr::rename(allele=V1, pct_ident=V3, start=V9, end=V10, positives_pct=V13, matchNaa=V4, query_length =V8) %>%
  mutate(pct_coverage = (matchNaa)/query_length*100
         ) %>%
  filter(pct_ident>80 & pct_coverage >80) %>%
  select(allele, start, end,pct_ident,pct_coverage,query_length)


p1a <- p1$gain_loss %>%
  dplyr::rename(strain=Var1, allele=Var2) %>%
  left_join(loc2, by='allele') %>%
  filter(value %in% c(-1,1) & !is.na(start)) %>%
  mutate(value=as.factor(value) ,strainN= as.numeric(as.factor(strain))+rnorm(n(), 0, 0.05) )

p2a <- ggplot(p1a, aes( y=strainN, x=start ,color=value,text=paste(strain,allele), alpha=0.5 )) +
         geom_point(size=0.3) +
       theme_classic() +
    xlim(range(loc2$start) ) +
  geom_vline(xintercept=c(56407, 2244856)) #roughly dexB and aliA

ggplotly(p2a)
```

