---
title: "Heatmaps of gene exchange experiment"
date: '2022-11-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(dplyr)
library(ggplot2)
library(plotly)
#library(d3heatmap)
library(viridis)
library(hrbrthemes)
library(heatmaply)
source('./R/plot.func.R')
```

## Credits
Analysis by Chrispin Chaguza
Experiments by Anna York
Sequencing by Jason Rosch and colleagues (St Jude's)
Interactive plots by Dan Weinberger

```{r}
key1 <-  read.table(file = './Data/Strain.clone.info.tsv', sep = '\t', header = TRUE) %>% rename(parent_ST=ID, new_ST=Strain , strain=Hartwell)

g1 <- read.table(file = './Data/Group1 gene.presence_absence.tsv', sep = '\t', header = TRUE)

g2 <- read.table(file = './Data/Group2 gene.presence_absence.tsv', sep = '\t', header = TRUE)

g3 <- read.table(file = './Data/Group3 gene.presence_absence.tsv', sep = '\t', header = TRUE)

```

## Serotype 9V background

Parent strain is Serotype 9V ID ABC010001831, obtained from the CDC

```{r, fig.width=10, fig.height=5}
plot.func(g1)

```

## Serotype 15B/C background

Parent strain is 15B/C, a carriage isolate from Hungary, ID Pool_hung_H10 (H124)

```{r, fig.width=10, fig.height=5}
plot.func(g2)

```

## Serotype 22F background

Parent strain is serotype 22F, ID: ABC010003966, obtained from the CDC (IPD isolate)

```{r, fig.width=10, fig.height=5}
plot.func(g3)

```




